---
title: "Gov 50 Final Project"
author: "Jama Willis"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Project Proposal
In Latin America and the Caribbean countries, does crime have an influence democratic institutions? This study conducts a bivariate analysis of the relationship between democratic institutions and crime within Latin America and the Caribbean. I hypothesize that crime impacts their democratic institutions and that crime has increased over the years therefore as crime increase, democracy in the Caribbean decreases. My data is compromised of the every countries V-Dem score indicators (all the factors that attribute to the overall democratic index of a country) from 1996 to 2021. My analysis will look at only Caribbean and Latin American countries. Although I am using this data for the purposes of my final project it is imperative to point out that the data collected are subject to several biases, including under representation of the developing world, asymmetries in reporting countries’ definitions of homicide, and those imposed by difficulties in collecting and compiling reliable data in nations with limited statistical and organizational capacity. The explanatory variable of interest is the democratic institutions or overall democratic index of each country and the outcome variable is the rate of crime rates in Latin Amercia and the Caribbean. 

```{r}
library(tidyverse)

dem_data <- read.csv("Democratic Index and RL Data Points - M3.csv")
dem_data <- dem_data |>
  select(EIU21RL, Democratic.Index..2021., Country)

ggplot(dem_data, aes(x=Democratic.Index..2021., y=EIU21RL))+
  geom_point(aes(color=Country)) +
  labs(x = "Democratic Index",
      y = "Rule of Law/Crime Index",
      title = "The Impact of Crime on Democratic Index in Latin American and Caribbean Nations")
```
The democratic index is the explanatory variable and the outcome variable is the rule of law or crime index. The plot suggests that there is a positive association between democratic index and rule of law/crime index, as defined by Economic Intelligence Unit. Democratic index measures based on five categories: electoral process and pluralism, civil liberties, the functioning of government, political participation, and political culture  and rule of law measures violent crime, organized crime, the fairness of judicial process, the enforceability of contracts, speediness of judicial process, confiscation/expropriation, intellectual property rights protection and private property protection. Thus, we see that countries that are considered more democratic generally are associated with stricter laws on crime. 

#Second Visualization

```{r}


dem_data <- read.csv("Democratic Index and RL Data Points - M3.csv")
dem_data <- dem_data |>
  select(EIU21RL, Democratic.Index..2021., Country)

ggplot(dem_data, aes(x=Democratic.Index..2021., y=EIU21RL))+
  geom_smooth(method = "lm", se=FALSE)+
  geom_point(aes(color=Country)) +
  labs(x = "Democratic Index",
      y = "Rule of Law/Crime Index",
      title = "The Impact of Crime on Democratic Index in Latin American and Caribbean Nations")
```
Based on this data, the line of regression shows a positive association between the Democratic index and the Rule of law/ Crime index. 


```{r}
dem_lm <- lm(EIU21RL ~ Democratic.Index..2021., data = dem_data)
dem_lm
```
The `lm` summary shows that there's a positive slope and that R^2 value shows that this is a relatively strong linear association. 
